---
title: "Project 1 markdown"
author: "cgibbs"
date: "Tuesday, October 14, 2014"
output: html_document
---
**Read the data**
```{r, echo=TRUE}
data<-read.csv("activity.csv")
```

**What is the mean total steps taken per day?**
1. Create a histogram of number of steps per day.
```{r,echo=TRUE} 
agg<-aggregate(data$steps, by=list(data$date), sum, na.rm=TRUE)

hist(agg$x, breaks=100,
     main = paste("Histogram of total number of steps per day"), 
     xlab= paste("Number of steps per day"))
```

2. Calculate the mean and median total steps per day
```{r, echo=TRUE} 
mean(agg$x)
median(agg$x)
```
**What is the average daily activity pattern?**

Make a time series plot (i.e. type = "l") 
of the 5-minute interval and the average number of steps
taken, averaged across all days.
```{r, echo=TRUE} 
agg.by.int<-aggregate(data$steps,
                      by=list(data$interval), 
                      mean, na.rm=TRUE)

plot(agg.by.int$Group.1, agg.by.int$x,
     type = "l", xlab= "Time Interval (minutes)", 
     ylab = "Number of Steps")
```

which 5 min interval has on average the maximum number of steps? 
*Interval 835 has the maximum average number of steps*

```{r, echo=TRUE}
agg.by.int[which.max(agg.by.int$x),]
```
**Imputing the data**

1.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
*There are 2304 rows with missing data.*

```{r, echo=TRUE}
missing<-is.na(data)
sum(missing)
```
2.Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

3.Create a new dataset that is equal to the original dataset but with the missing data filled in. 

*Using the dplyr package, the mean for the given 5-minute time interval was used to replace all missing values*
```{r, echo=TRUE}
library(dplyr)

data.imputed<-data %>% 
    group_by(interval) %>% 
    mutate(steps= ifelse(is.na(steps), mean(steps, na.rm=TRUE), 
    steps))

head(data.imputed)
```

4.Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps? 

*Values for mean and median differ before and after imputation. Both mean and median values are higher after imputation at 10,766.19, and the median is exactly equal to the mean now.* 

```{r, echo=TRUE}
agg.steps.imputed<-aggregate(data.imputed$steps, 
                  by = list(data.imputed$date), sum)

hist(agg.steps.imputed$x, breaks=100, 
     main = paste("Histogram of total number of steps per day with imputed data"), 
     xlab= paste("Number of steps per day"))

mean(agg.steps.imputed$x)
median(agg.steps.imputed$x)
```
**Are there differences in activity patterns between weekdays and weekends?**

1.First, create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r, echo=TRUE}
data$day<-weekdays(as.Date(data$date))
data$day[data$day == "Monday"]<-"weekday"
data$day[data$day == "Tuesday"]<-"weekday"
data$day[data$day == "Wednesday"]<-"weekday"
data$day[data$day == "Thursday"]<-"weekday"
data$day[data$day == "Friday"]<-"weekday"
data$day[data$day == "Saturday"]<-"weekend"
data$day[data$day == "Sunday"]<-"weekend"
data$day<-as.factor(data$day)
```

2.Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was creating using simulated data:

```{r, warning=FALSE, echo=TRUE, fig.width = 9, fig.height=4}
library(ggplot2)

p<-ggplot(data, aes(interval,steps,group=day))+
  stat_summary(fun.y="mean",geom="line",color="blue")+
  facet_wrap(~day) + xlab("Time Interval (minutes)") + ylab("Average Number of Steps")

p
```

*Several differences appear in these two plots. First the number of steps increases at an earlier time interval on weekdays, but also, there is a is a large spike in number of steps (206.1 steps) in the 500-1000 interval range and decline towards 0 before the 2000 time interval.  On weekends, the number of steps increases beginning at a later interval, but does not reach that peak shown on the weekday plot, and the number of steps decline at a later time interval, after the 2000 interval range.*